---
title: "Seasonal Microbe Effect on Duckweed"
author: "Emma Kinerson"
format: pdf
fig_caption: true
editor: visual
fontsize: 12pt
bibliography: "806_Final_references.bib"
---
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(readxl)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(car)
library(lmerTest)
library(broom.mixed)
library(webshot)
library(htmlwidgets)
library(kableExtra)
library(knitr)
library(ggsignif)
library(emmeans)
library(ggpubr)


my_path <- "C:/Users/emmae/Documents/Undergrad Plate Data- OBrien_combined.xlsx"
data <- read_excel(my_path)
```

```{r, echo=FALSE}
long_data <- data %>%
  pivot_longer(cols = `1`:`12`,            # Select all columns with OD values (1-12)
               names_to = "Column",        # Create a new column for the column number
               values_to = "OD_values") %>%  # New column for OD values
  mutate(Well_ID = paste(...2, Column, sep = ""))  # Combine row and column to form Well_ID
```

```{r, echo=FALSE}
my_path_2 <- "C:/Users/emmae/Documents/Plate_setup_undergrad_data_Obrien.xlsx"
plate_setup <- read_excel(my_path_2)

row_mapping <- c("1" = "A", "2" = "B", "3" = "C", "4" = "D", 
                 "5" = "E", "6" = "F", "7" = "G", "8" = "H")
plate_setup <- plate_setup %>% 
  mutate(
    Row = row_mapping[as.character(Row)],  # Convert Row numbers to letters using mapping
    Well_ID = paste0(Row, Col)          # Concatenate Row and Column to form Well_ID
  )

plate_setup <- plate_setup %>% 
   rename(`Plate #` = Plate)

```

```{r, echo=FALSE}
combined_data <- left_join(long_data, plate_setup, by = c("Well_ID", "Plate #"))

combined_data$...2 <- NULL
combined_data$Col <- NULL
```

```{r, echo=FALSE}
#getting rid of the unknown blanks
combined_data_no_water <- combined_data %>%
  filter(Micr != "Water", Plant != "Water")
```

```{r, echo=FALSE}
#Getting Rid of Outliers Read 1
threshold <- 1.0  # Set your threshold value

# Filter out rows where OD is greater than the threshold
filtered_combined_NW_1 <- combined_data_no_water %>%
  filter(OD_values <= threshold, Read == "Read 1:450")
```

```{r echo=FALSE}
#Getting Rid of Outliers Read 2
threshold <- 1.0  # Set your threshold value

# Filter out rows where OD is greater than the threshold
filtered_combined_NW_2 <- combined_data_no_water %>%
  filter(OD_values <= threshold, Read == "Read 2:600")
```



**Introduction:**

  The complex interaction that occurs between microorganisms and plants plays an important role in supporting the health and functionality of multicellular organisms (@obrien_mutualistic_2020). In plants, microbial communities significantly enhance fitness through improving nutrient uptake, enhancing stress tolerance, and protecting against pathogens (@obrien_mutualistic_2020). These relationships are central to ecosystem stability and biodiversity preservation. The benefits of these interactions, are influenced by various environmental factors, including temperature, soil composition, and seasonal changes (@ruan_elevated_2023; @santoyo_rhizosphere_2021). As climate change continues to reshape Earth’s ecosystems, understanding these interactions has become increasingly critical for developing strategies to maintain biodiversity as well as agricultural productivity. 
  
  Seasonal changes in particular, can cause shifts in both the composition and functionality of microbial communities. For example, Ma et al. (2021) found that soil microbiomes surrounding hazelnut plants varied across seasons which in turn altered nutrient availability for the host plants (@ma_seasonal_2021). While it is understood that environmental conditions can reshape microbial communities, the effects of these shifts on the plant hosts remain unclear. It is important to understand these effects and how the overall dynamic changes to be able to predict plant-microbe interactions in fluctuating environments to be able to improve plant fitness. 

  To address this knowledge gap, duckweed (*Lemna minor*) was used as a model organism. Duckweed is a small, free-floating aquatic plant with a rapid growth rate and short life cycle. These characteristics make it an ideal model for studying generational changes and effects of environmental variables (@thingujam_duckweed_2024-1). Because duckweed is sensitive to environmental fluctuations, it provides an excellent system for investigating how seasonal shifts in microbiomes impact plant growth and fitness (@thingujam_duckweed_2024-1). This study hypothesized that microbiomes collected during different seasons will vary in their growth as well as their impact on duckweed growth, as measured by plant area and optical density of microbial communities, reflecting the seasonal shifts in microbial community composition. 

**Methods:**

1.  Microbe Collection:

  Microbes were sampled from two different locations across all four seasons. Seven sample collections were completed at the La Roche sample site, and eight sample collections were completed at the Thompson Farm site. Samples were collected at the edge of a body of water using sterilized metal spatulas . Collection tubes were rinsed in the body of water three times before being used to collect duckweed and the surrounding water.
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Map showing the locations of Thompson Farm and La Roche sampling areas. Both locations are located Southwest of the University of New Hampshires campus.", fig.pos = "!h"}
library(leaflet)

 map <- leaflet() %>% 
  addTiles() %>% 
  # Add marker for Thompson Farm with a label
  addLabelOnlyMarkers(lng = -70.94978753239516, lat = 43.1104921981106770,
                      label = "Thompson Farm", 
                      labelOptions = labelOptions(noHide = TRUE, direction = "top")) %>%
  # Add marker for La Roche with a label
  addLabelOnlyMarkers(lng =  -70.9460138783696, lat = 43.11948257344687,
                      label = "La Roche", 
                      labelOptions = labelOptions(noHide = TRUE, direction = "top")) %>% 
 setView(lng = -70.947, lat = 43.115, zoom = 14) 
 
saveWidget(map, "map.html")
 webshot("map.html", file = "map_screenshot.png", vwidth = 800, vheight = 600)

```

2.  Duckweed Line:

  Isogenic duckweed lines were created using duckweed from each respective site. Single fronds from each site were placed into twenty four well plates and allowed to asexually reproduce. From these two lines, axenic cultures were created for each respective line. These cultures were created by sterilizing the duckweed using a 1% bleach solution for one minute and thirty seconds. Following the bleach sterilization, the plants were rinsed three times with sterile water. The first rinse lasted forty five seconds with the following two rinses lasting ten minutes each. The sterile duckweed was then placed into sterile Krazčič media (K media) for each respective isogenic line. The duckweed within each new axenic culture were then plated on sterile enriched yeast mannitol media to test for microbe growth to ensure sterilization.

3.  Culturing Microbes:

  Duckweed from the microbe samples were crushed on the inside cover of a plate that contained yeast mannitol agarose using a sterilized inoculation loop. This inoculation loop was then used to streak the plate to culture the microbes from the duckweed. These plates were then incubated until colony growth was sufficient. Once sufficient growth was achieved, they were stored in a 4ºC refrigerator until use.

4.  Experiment Design and Set Up:

  Both axenic duckweed lines were inoculated with each seasonal collection of microbes from both duckweed lines respective sample site. These inoculations were set up with 4 blocks containing 8 replicates each into six 96 well plates according to a randomized blocking map. This allowed for 32 replicates for each treatment. To create the experimental inocula from the microbe samples, 15 mL culture tubes containing sterile yeast mannitol media were inoculated by swabbing with microbes from each combination of site and date. The experimental combinations of these microbes were cultured for two days at an RPM of 300 and at 30ºC. A BioTek Cytation 5 Plate Reader was used to examine optical density at 600 nm for each microbe inocula to test cell concentration (@obrien_mutualistic_2020). Cell concentrations from the OD readings wree used to dilute all inocula to the same concentration (5000 cells/µl). Dilutions were formed by adding a calculated amount of microbe inocula to sterile Hogland's media in the laminar flow hood. Well plates were prepped by placing 600 µl of sterilized Hogland's media into each well, followed by placing duckweed fronds in each well. A seal (BreathEasier Diversified Biotech) was placed onto each plate and then the plates were left in the sterile laminar flow hood overnight. Each microbe inocula (10 µl) were added to the 96 well plates according to the randomized block design. This was completed by puncturing through the Breatheasier seal. Once all wells were inoculated, the BreathEasier seal was removed and a new seal (BreathEasy Diversified Biotech) was placed onto the plate to prevent contamination. Plates were placed into the growth chamber and imaged on day 12. Images were taken using an iPhone 13 cell phone camera. Images were taken through the plexiglass shield that the plate were set upon to ensure uniform images. ImageJ was used for analysis to measure duckweed growth.
    
5. Post Microbe Analysis: 

  After the treatment period, the media containing the microbial samples was analyzed for optical density (OD) using a microplate reader (BioTek Cytation 5 Plate Reader) to assess microbial growth. Samples were transferred from experimental plates into a clear bottom 96-well microplate. The optical density was measured at two different wavelengths: 450 nm (OD450) and 600 nm (OD600), to capture both general microbial growth and specific cell concentration. 

6.  Statistical Tests:

  Statistical tests were conducted after the experiment to analyze the data. Both OD450 and OD600 wavelengths were initially considered for analysis. A Spearman rank correlation test was conducted to assess the relationship between the two wavelengths, and a strong positive correlation was observed (rho = 0.85, p < 0.001). Based on this high degree of similarity between the two measurements, and given that OD600 is commonly used to assess microbial growth, it was selected for further analysis as it directly reflects the microbial biomass in the samples (@mira_estimating_2022). This decision was made to align with standard practice in microbial studies and to ensure the relevance of the measurements to the research goals.
 
  To account for the random effects of replicates, a linear mixed model (LMM) was then applied to the log-transformed optical density data. The log transformation was applied based on model fit, as indicated by the AIC criterion. The LMM was used to compare the effects of seasonal microbial treatments against the control group (which contained no microbes) and to evaluate differences between the two plant locations. While the assumption of normality was slightly violated, this was deemed acceptable due to the large sample size (greater than 500 samples), as supported by the Central Limit Theorem. Additionally, a post hoc Tukey test was performed to assess pairwise comparisons between the groups. Given the large sample size, the slight departure from normality was considered negligible, and the results of the Tukey test were interpreted accordingly.


**Results:**

Relationship Between Microbial OD450 and OD600 Measurements:

To visualize the correlation between microbial optical density (OD) readings at 450 nm and 600 nm, a scatter plot was generated (Figure 2), accompanied by a Spearman's rank correlation test (Table 1). The scatter plot showed a positive correlation between the reading at 450 nm and the reading at 600 nm. This positive correlation suggests that as OD values increase at the 450 nm wavelength that they also increase in the 600 nm wavelength read. The Spearman’s correlation coefficient (ρ = 0.98) indicated a strong correlation between the two read wavelengths. However, a p-value could not be generated in the test due to ties in the data set, which can be caused by identical values in the data. Despite this the OD450 and OD600 values were generally consistent and the strong correlation supports the decision to move forward with OD600 values for further analysis. 

```{r, echo=FALSE, warning=FALSE, fig.cap="Correlation between OD450 and OD600 readings across all samples. The strong positive correlation (visualized by the red regression line) indicates consistent measurement trends between the two wavelengths", fig.pos = "!h"}

combined_data_NW_NOut_wider <- combined_data_no_water %>%
  filter(OD_values <= threshold) %>% 
  pivot_wider(
    names_from = Read,  # This will use the Wavelength column for new column names
    values_from = OD_values)   # This will fill the new columns with the corresponding OD values

# Scatterplot for visualizing the relationship between OD450 and OD600
plot(combined_data_NW_NOut_wider$`Read 1:450`, combined_data_NW_NOut_wider$`Read 2:600`, main = "OD450 vs OD600", xlab = "OD450", ylab = "OD600")
abline(lm(`Read 1:450` ~ `Read 2:600`, data = combined_data_NW_NOut_wider), col = "red") # Add regression line
```

```{r, echo=FALSE, warning=FALSE, fig.pos = "!h"}
cor_test_result <- cor.test(combined_data_NW_NOut_wider$`Read 1:450`, combined_data_NW_NOut_wider$`Read 2:600`, method = "spearman", conf.level = 0.95)

# Extract relevant values from correlation test result
cor_coefficient <- cor_test_result$estimate  # Spearman's rho
p_value <- cor_test_result$p.value
cor_statistic <- cor_test_result$statistic
method <- cor_test_result$method

# Create a table of correlation test results with separate rows for confidence interval
results_table <- data.frame(
  Statistic = c("Correlation Coefficient", "p-value",
                "Test Statistic", "Method"),
  Value = c(cor_coefficient, p_value, cor_statistic, method)
)

results_table <- results_table %>%
  mutate(across(where(is.numeric), ~ signif(.x, 3)))

# Print the table using kable
kable(results_table, caption = "Spearman's Rank Correlation Test Results between Variables OD450 and OD600") %>% 
   kable_styling(latex_options = "hold_position")
```
Seasonal and Plant Location Variation in Microbe OD Value:

The mean log-transformed OD600 values were plotted for each season (Figure 3). The values were log-transformed due to AIC value (AIC = 104) determining that it was the best fit for the data. The control with no microbes exhibited the lowest mean OD600 value, with seasonal microbe treatments displaying more variation. Summer and winter treatments appear to have slightly higher mean OD values, with winter displaying the greatest variability. This suggests differences across seasons in microbial growth or activity. 

```{r, echo=FALSE, fig.cap= "Mean log-transformed optical density (OD) values of microbial growth across different seasons, including a control with no microbes. Error bars represent the standard deviation. Seasonal microbial communities were analyzed for their effect on optical density, highlighting variations in microbial growth between seasons.", fig.pos = "!h"}

season_mean_summary <- filtered_combined_NW_2 %>% 
  group_by(Season) %>% 
  summarise(
    mean_log_OD = mean(log(OD_values)),
    sd_log_OD = sd(log(OD_values)),
    min_log_OD = min(log(OD_values)),
    max_log_OD = max(log(OD_values))
  )

ggplot() +
  geom_point(data = season_mean_summary, aes(x = Season, y = mean_log_OD, color = Season), 
             shape = 18, size = 4) +
  geom_errorbar(data = season_mean_summary, aes(x = Season, ymin = mean_log_OD - sd_log_OD, ymax = mean_log_OD + sd_log_OD), 
                width = 0.2)  +
  labs(title = "Mean Log Opitcal Density Values by Season",
       x = "Season",
       y = "Log OD Values",
       color = "Season") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")

```
A heat map displayed in Figure 4, compares OD600 values of microbe samples by both season and plant location providing a more nuanced perspective into differences in values (Figure 4). Patterns between plant location and season emerged. There was a clear visual difference between OD600 values for certain plant-season combinations, with higher OD600 values being present in microbes that were associated with plants from the LaRoche sample site. Comparisons of microbes across seasons from the sampling site at LaRoche visually depict slightly higher OD600 values in spring and summer. This continues to suggest seasonal differences in microbial growth or activity. This also suggests that plant location could play a role in microbial composition and its susceptibility to environmental changes such as seasons. 

```{r, echo=FALSE, fig.cap= "Heatmap of microbe log-transformed optical density by both season and plant location. Lighter colors indicate a lower optical density, with darker colors representing a higher optical density value.", fig.pos = "!h"}
filtered_combined_NW_2$log_OD_values <- log(filtered_combined_NW_2$OD_values)

ggplot(filtered_combined_NW_2, aes(x = Plant, y = Season, fill = log_OD_values)) +
  geom_tile(color = "black", aes(fill = log_OD_values)) +
  scale_fill_gradient2(low = "white", mid = "darkblue", high = "red", midpoint = 0, 
                      name = "Negative Log Optical Density") +
  theme_minimal() +
  labs(title = "Heatmap of Negative Log-Transformed Optical Density by Season and Plant Location",
       x = "Plant Location", y = "Season")
```
Linear Mixed Model (LMM) Results:

To quantify the effects of season as well as plant location on OD600 values from microbial samples while accounting for random variation among replicates, a linear mixed model (LMM) was applied (Table 2). The OD values were log transformed before the LMM was applied, as an AIC value of 104 confirmed that it was a good fit for the model. The model revealed mutliple significant fixed effects for season. Fall (p = 0.041), winter (p = 0.00049), and summer (p = 0.00005) were all significantly different from the intercept which was the control condition. The model also revealed that the plant locations were significantly different (p = 0.0024). This model suggests that both season as well as plant location impact microbial growth.
```{r, echo=FALSE, warning=FALSE, fig.pos = "!h"}
# Load necessary libraries
library(lme4)
library(broom.mixed)

# Apply log transformation to the OD values
filtered_combined_NW_2$log_OD_values <- log(filtered_combined_NW_2$OD_values)

# Run the linear mixed model with the log-transformed data
lmm_result_log_read2 <- lmer(log_OD_values ~ Season + Plant + (1| rep), data = filtered_combined_NW_2)


# Extract fixed effects summary
lmm_summary_read2 <- summary(lmm_result_log_read2)$coefficients
fixed_effects <- as.data.frame(lmm_summary_read2)

# Rename columns for clarity
fixed_effects <- fixed_effects %>%
  rename(
    `Estimate` = Estimate,
    `Standard Error` = `Std. Error`,
    `Degrees of Freedom` = df,
    `t Value` = `t value`,
    `p Value` = `Pr(>|t|)`
  )

# Round numeric values for better presentation
fixed_effects <- fixed_effects %>%
  mutate(across(where(is.numeric), ~ signif(.x, 3)))

# Create the table using knitr::kable()
kable(fixed_effects, 
      caption = "Fixed Effects of the Linear Mixed Model",
      format = "latex",  # Use "html" or "latex" for other output formats
      align = "c") %>%
  kable_styling(full_width = FALSE, position = "center", latex_options = "hold_position")

```

Pairwise Comparisons Using Tukey Adjustments:

Post hoc comparisions were completed to determine differences in OD600 values across seasons and plant location (Table 3, Table 4). There were two significant differences for seasons. The log-transformed OD was significantly lower in the control compared to summer, with an estimated mean difference of -0.151 (SE = 0.037, df = 532, t = -4.09, p = 0.0005). The control was also significantly lower than winter with an estimated mean difference of -0.159 (SE = 0.045, df = 532, t = -3.51, p = 0.0044). When compared to fall and spring there were no significant differences. Between seasons, the only significant comparison was between spring and summer with spring having significantly lower OD values based on an estimated mean difference of -0.107 (SE = 0.037, df = 532, t = -2.9, p = 0.032). The post hoc Tukey test revealed a significant difference in the log-transformed OD values between the two plant locations, LaRoche and Thompson Farm (Table 4). The estimated mean difference was 0.0673 (SE = 0.022, df = 532, t = 3.06, p = 0.002). This indicates that microbial communities collected from plants at the LaRoche site had higher OD values compared to those from Thompson Farm. 

```{r, echo=FALSE, warning=FALSE, fig.pos = "!h"}
# Calculate estimated marginal means (EMMs)
emm_season_read2 <- emmeans(lmm_result_log_read2, ~ Season)

# Perform pairwise comparisons with Tukey adjustment
tukey_results_season_read2 <- pairs(emm_season_read2, adjust = "tukey")


# Convert the results to a data frame
tukey_results_df <- as.data.frame(tukey_results_season_read2)

# Rename columns for clarity
tukey_results_df <- tukey_results_df %>%
  rename(
    `Comparison` = contrast,
    `Estimate` = estimate,
    `Standard Error` = SE,
    `t Ratio` = t.ratio,
    `p Value` = p.value
  )

# Round numeric values for better presentation
tukey_results_df <- tukey_results_df %>%
  mutate(across(where(is.numeric), ~ signif(.x, 3)))

# Create the table using knitr::kable()
kable(tukey_results_df, 
      caption = "Tukey Post Hoc Pairwise Comparisons for Season (Estimated Marginal Means)",
      format = "latex",  # Use "html" or "latex" depending on the output format
      align = "c") %>%
  kable_styling(full_width = FALSE, position = "center", latex_options = "hold_position")
```


```{r, echo=FALSE, warning=FALSE, fig.pos = "!h"}
# Calculate estimated marginal means (EMMs)
emm_plant_read2 <- emmeans(lmm_result_log_read2, ~ Plant)

# Perform pairwise comparisons with Tukey adjustment
tukey_results_plant_read2 <- pairs(emm_plant_read2, adjust = "tukey")


# Convert the results to a data frame
tukey_results_df_plant <- as.data.frame(tukey_results_plant_read2)

# Rename columns for clarity
tukey_results_df_plant <- tukey_results_df_plant %>%
  rename(
    `Comparison` = contrast,
    `Estimate` = estimate,
    `Standard Error` = SE,
    `t Ratio` = t.ratio,
    `p Value` = p.value
  )

# Round numeric values for better presentation
tukey_results_df_plant <- tukey_results_df_plant %>%
  mutate(across(where(is.numeric), ~ signif(.x, 3)))

# Create the table using knitr::kable()
kable(tukey_results_df_plant, 
      caption = "Tukey Post Hoc Pairwise Comparisons for Plant (Estimated Marginal Means)",
      format = "latex",  # Use "html" or "latex" depending on the output format
      align = "c") %>%
  kable_styling(full_width = FALSE, position = "center", latex_options = "hold_position")

```


\pagebreak
**Discussion:**

The study evaluated microbial impacts on duckweed growth across seasons, with data analysis conducted so far focusing on efect of season and plant location on microbial growth post experimental exposure with duckweed. OD600 measurements showed strong correlations with OD450 measurements. Due to the positive correlations between the two measures as well as previous studies using OD600 measurements confirmed its reliability for further analysis (@obrien_mutualistic_2020). The further analysis observed statistically significant growth in the microbial sample for the summer months. This may be due to increased microbial activity during favorable environmental conditions. There is also a potential of higher diversity within the community from the summer months that could cause enhanced microbial growth. The lack of significance for the spring months could potentially be explained by unfavorable environmental conditions transitioning from winter to spring. Overall, the initial findings align with prior research that shows seasonal variations influence functions of microbial communities (@santoyo_rhizosphere_2021). 

The largest limitation of the analysis within this study was the slight non-normality present within the optical density data which could potentially affect the results of the statistical models. However, the large sample size (n > 500) helpts to mitigate the error this limitation could cause under the central limit theorem. Additionally, there were tied/identical OD values within the data set which may have impacted the results of the Spearman's correlation test. Another notable limitation was the high variability observed in OD values for the controls, which could be attributed to incomplete plant sterilization, leaving residual microbes that elevated OD values.

Further analysis will be completed to analyze the impacts of these microbe samples from the various combinations of seasons and plant location on duckweed growth. To analyze the impact on growth the mean frond area will be compared between the control and experimental conditions. The results from this complementary analysis will provide a holistic understanding of how seasonal microbes and plant location influence plant growth. To further contextualize these results, microbial sequencing data will also be added into the analysis to understand the differences in community composition across seasons and how that correlates with plant growth outcomes. Future studies to further the understanding of these dynamics as well as their application should examine the long term stability  and consistency of beneficial plant growth promoting bacterial communities across seasons. 

This study contributes to understanding the role of environmental factors on microbial communities and, in turn, their effects on plant growth. Seasonal variability in microbial efficacy can help to enhance bio-remediation strategies and agricultural applications. Insights gained from this research can help to inform microbial community selection in aquatic systems, with potential extension to soil applications. By enhancing knowledge of how environmental factors influence microbial functionality, this study helps to form the foundation for future innovations in sustainable agricultural practices. 

\pagebreak

**References:**










